---
title: "02 - Exploratory Plots"
author: "Lina Walkowiak"
date: "`r Sys.Date()`"
output: html_document
---
# Exploration plots
## Load packages and data 
```{r}
library(sf)
library(tmap)
library(dplyr)
library(readr)
library(tidyr)
library(ggplot2)

# Load preprocessed data
census_sections <- readRDS("../data/processed/census_sections.rds")
combined <- readRDS("../data/processed/both_years.rds")
before <- readRDS("../data/processed/before_final.rds")
after <- readRDS("../data/processed/after_final.rds")




```

## Merge data and geometry 
```{r}
before_sf <- census_sections %>%
  left_join(st_drop_geometry(before), by = "census_id")

after_sf <- census_sections %>%
  left_join(st_drop_geometry(after), by = "census_id")

combined_sf <- census_sections %>%
  left_join(st_drop_geometry(combined), by = c("census_id" = "census_id"))

```

### Airbnb data
```{r}

# -------- 2015 ------------
ggplot() +
  # Census sections
  geom_sf(data = census_sections, fill = "grey95", color = "grey70", size = 0.4) +
  
  # Airbnb listings
  geom_sf(data = listings2015, aes(color = room_type), alpha = 0.5, size = 1.4) +

  # Custom color palette
  scale_color_manual(values = c("Entire" = "#1f78b4", "Room" = "#33a02c", "Other" = "#ff7f00")) +

  labs(
    title = "Airbnb Listings by accomodation type",
    subtitle = "Barcelona (2015)",
    color = "Accomodation Type"
  ) +

  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 11, hjust = 0.5),
    legend.position = "bottom",
    panel.grid = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank()
  )



# ---------- 2025 -------------
ggplot() +
  # Census sections
  geom_sf(data = census_sections, fill = "grey95", color = "grey70", size = 0.4) +
  
  # Airbnb listings
  geom_sf(data = listings, aes(color = room_type), alpha = 0.5, size = 1.4) +

  # Custom color palette
  scale_color_manual(values = c("Entire" = "#1f78b4", "Room" = "#33a02c", "Other" = "#ff7f00")) +

  labs(
    title = "Airbnb Listings by accomodation type",
    subtitle = "Barcelona (2025)",
    color = "Accomodation Type"
  ) +

  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 11, hjust = 0.5),
    legend.position = "bottom",
    panel.grid = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank()
  )
```

## Delta variabels 
```{r}
library(tmap)
tmap_mode("plot")
```

### income 
```{r}
tm_shape(combined_sf) +
  tm_polygons("delta_income",
              palette = "YlOrRd",
              title = "Δ Income (€)",
              style = "quantile",
              border.col = NA) +
  tm_layout(
    title = "Change in Disposable Income (2015–2021)",
    legend.outside = TRUE,
    legend.frame = FALSE,          # Removes legend border
    legend.position = c("right", "bottom"), 
    frame = FALSE                  # Removes map frame
  )




```


### education level
```{r}
tm_shape(combined_sf) +
  tm_polygons("delta_pct_edu_5",
              palette = "YlOrRd",
              title = "Δ % Highly Educated",
              style = "quantile",
              border.alpha = 0.1) +
  tm_layout(title = "Change in share of highly educated residents",
               legend.outside = TRUE,
               legend.frame = FALSE,          # Removes legend border
               legend.position = c("right", "bottom"), 
               frame = FALSE                  # Removes map frame
  )

# Different colour 
tm_shape(combined_sf) +
  tm_polygons("delta_income", palette = "RdYlBu", title = "Δ Income (2021 - 2015)", style = "quantile") +
  tm_layout(title = "Change in Disposable Income")


```



```{r}

library(leaflet)
library(sf)
library(dplyr)

# 1. Transform to WGS84 (required for leaflet)
combined_wgs <- st_transform(combined_sf, crs = 4326)

# 2. Define global range for color scaling (shared across levels)
global_range <- range(c(combined_wgs$delta_pct_edu_1,
                        combined_wgs$delta_pct_edu_2,
                        combined_wgs$delta_pct_edu_3,
                        combined_wgs$delta_pct_edu_4,
                        combined_wgs$delta_pct_edu_5,
                        combined_wgs$delta_pct_edu_6), na.rm = TRUE)

# 3. Create color palette
pal <- colorNumeric("YlOrRd", domain = global_range, na.color = "transparent")


leaflet(data = combined_wgs) %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  
  addPolygons(fillColor = ~pal(delta_pct_edu_1),
              color = "#444", weight = 0.5, fillOpacity = 0.7,
              label = ~paste0("Δ Level 1: ", round(delta_pct_edu_1 * 100, 2), "%"),
              group = "Level 1") %>%
  
  addPolygons(fillColor = ~pal(delta_pct_edu_2),
              color = "#444", weight = 0.5, fillOpacity = 0.7,
              label = ~paste0("Δ Level 2: ", round(delta_pct_edu_2 * 100, 2), "%"),
              group = "Level 2") %>%
  
  addPolygons(fillColor = ~pal(delta_pct_edu_3),
              color = "#444", weight = 0.5, fillOpacity = 0.7,
              label = ~paste0("Δ Level 3: ", round(delta_pct_edu_3 * 100, 2), "%"),
              group = "Level 3") %>%
  
  addPolygons(fillColor = ~pal(delta_pct_edu_4),
              color = "#444", weight = 0.5, fillOpacity = 0.7,
              label = ~paste0("Δ Level 4: ", round(delta_pct_edu_4 * 100, 2), "%"),
              group = "Level 4") %>%
  
  addPolygons(fillColor = ~pal(delta_pct_edu_5),
              color = "#444", weight = 0.5, fillOpacity = 0.7,
              label = ~paste0("Δ Level 5: ", round(delta_pct_edu_5 * 100, 2), "%"),
              group = "Level 5") %>%
  
  addPolygons(fillColor = ~pal(delta_pct_edu_6),
              color = "#444", weight = 0.5, fillOpacity = 0.7,
              label = ~paste0("Δ Level 6: ", round(delta_pct_edu_6 * 100, 2), "%"),
              group = "Level 6") %>%
  
  addLayersControl(
    baseGroups = c("Level 1", "Level 2", "Level 3", "Level 4", "Level 5", "Level 6"),
    options = layersControlOptions(collapsed = FALSE)
  ) %>%
  
  addLegend(pal = pal,
            values = global_range,
            title = "Δ Share (%)",
            position = "bottomright")


```


#### Description of eductaion levels 
| Level | Likely Meaning                           |
| ----- | ---------------------------------------- |
| 1     | No formal education / Illiterate         |
| 2     | Primary education                        |
| 3     | Lower secondary (ESO)                    |
| 4     | Upper secondary (Bachillerato or FP)     |
| 5     | Tertiary education (University or FP II) |
| 6     | Postgraduate (Masters, PhD, etc.)        |


### ownership
```{r}
tm_shape(combined_sf) +
  tm_polygons("delta_pct_legal_entity",
              palette = "YlOrRd",
              title = "Δ % Corporate Ownership",
              style = "quantile",
              border.alpha = 0.1) +
  tm_layout(title = "Change in Legal Entity Ownership (2018–2025)",
    legend.outside = TRUE,
    legend.frame = FALSE,          # Removes legend border
    legend.position = c("right", "bottom"), 
    frame = FALSE                  # Removes map frame
  )

# different colour 
tm_shape(combined_sf) +
  tm_polygons("delta_pct_legal_entity",
              palette = "inferno", style = "quantile",
              title = "Δ Corporate Ownership") +
  tm_layout(title = "Change in Corporate Property Ownership (2025 – 2018)",
            legend.outside = TRUE)

```

## for paper 
```{r}
library(tmap)

tmap_mode("plot")  # or "view" for interactive map

tm_shape(before_sf) +
  tm_polygons(
    col = "Import_Euros",
    fill.scale = tm_scale_intervals(
      style = "quantile",   # can change to "jenks", "equal", or use breaks=
      n = 5,
      values = "YlOrRd"
    ),
    fill.legend = tm_legend(title = "Income (€)"),
    border.col = "white",
    col_alpha = 0.8
  ) +
  tm_title("Income Distribution by Census Section (2015)") +
  tm_layout(
    legend.outside = TRUE,
    legend.position = c("right", "bottom"),
    legend.frame = FALSE,
    frame = FALSE
  )



```




