---
title: "03 - Kernel-Based Spillover"
author: "Lina Walkowiak"
date: "`r Sys.Date()`"
output: html_document
---

### Kernel-Based Spillover from Airbnb Listings (Outside Section)

In this script, we compute a **spillover metric** for each census section that measures the intensity of Airbnb presence *just outside* its borders to account for the proximity of airbnbs that are not included in teh density measure. The idea is that Airbnb activity nearby (but not within) a section can still influence its socio-economic development, e.g., through housing pressure, tourist behavior, or property investment.

- Only include listings **outside** the section.
- Use the **distance to the section's border** (not centroid here?).
- Apply a **kernel function** to weight each nearby listing based on distance (closer = stronger influence)

### Load packages and data
```{r}
library(sf)
library(dplyr)
library(geosphere)
library(purrr)
library(readr)

# Load spatial objects
census_sections <- readRDS("../data/processed/census_sections.rds")
listings2015_sf <- readRDS("../data/processed/listings2015_sf.rds")
listings2025_sf <- readRDS("../data/processed/listings_sf.rds")

### 
projected_crs <- 25831  # Use a projected one for distance-based calculations
census_sections <- st_transform(census_sections, crs = projected_crs)
listings2015_sf <- st_transform(listings2015_sf, crs = projected_crs)
listings2025_sf <- st_transform(listings2025_sf, crs = projected_crs)


# Read previous internal mapping of listings
listings2015_within <- st_join(listings2015_sf, census_sections["census_id"], left = FALSE)
listings2025_within <- st_join(listings2025_sf, census_sections["census_id"], left = FALSE)
```

### Prepare border geometries for all sections
```{r}
census_edges <- census_sections %>%
  select(census_id) %>%
  st_boundary()
```

### Define kernel weighting function
```{r}
kernel_weight <- function(distance, sigma = 300) {
  exp(-distance / sigma)  # Exponential decay kernel? we can try different things here
}
```

### Function to compute kernel spillover for one section
```{r}
compute_kernel_spillover <- function(target_id, borders, listings, listings_within, sigma = 300) {
  border_geom <- borders %>% filter(census_id == target_id)

  # Listings not in this census section
  listings_outside <- listings %>%
    filter(!(id %in% listings_within$id[listings_within$census_id == target_id]))

  if (nrow(listings_outside) == 0) {
    return(tibble(census_id = target_id, kernel_spillover = 0))
  }

  # Distance from each external listing to border
  distances <- st_distance(listings_outside, border_geom)
  distances_m <- as.numeric(distances)  # Convert units here

  weights <- kernel_weight(distances_m, sigma)
  spillover_sum <- sum(weights, na.rm = TRUE)

  tibble(census_id = target_id, kernel_spillover = spillover_sum)
}
```

### Apply to all sections
```{r}
sigmas <- c(100, 300, 500)

kernel_spillovers_2015 <- map_dfr(
  sigmas,
  function(s) {
    map_dfr(
      census_sections$census_id,
      function(c_id) {
        out <- compute_kernel_spillover(
          c_id,
          borders = census_edges,
          listings = listings2015_sf,
          listings_within = listings2015_within,
          sigma = s
        )
        # Attach correct census_id manually
        out$census_id <- c_id
        out
      }
    ) %>%
      mutate(sigma = s)
  }
)


# for 2025

kernel_spillovers_2025 <- map_dfr(
  sigmas,
  function(s) {
    map_dfr(
      census_sections$census_id,
      function(c_id) {
        out <- compute_kernel_spillover(
          c_id,
          borders = census_edges,
          listings = listings2025_sf,
          listings_within = listings2025_within,
          sigma = s
        )
        # Attach correct census_id manually
        out$census_id <- c_id
        out
      }
    ) %>%
      mutate(sigma = s)
  }
)



# pivot to wide
spillover_all_2015 <-  kernel_spillovers_2015 %>%
  mutate(sigma = paste0("kernel_spill_", sigma)) %>%
  pivot_wider(names_from = sigma, values_from = kernel_spillover)

spillover_all_2025 <-  kernel_spillovers_2025 %>%
  mutate(sigma = paste0("kernel_spill_", sigma)) %>%
  pivot_wider(names_from = sigma, values_from = kernel_spillover)

```

### Join to main dataset
```{r}
full_2015 <- readRDS("../data/processed/before_processed.rds")
before_final <- left_join(spillover_all_2015, full_2015, by = "census_id")
saveRDS(before_final, "../data/processed/before_final.rds")


# same for later year
full_2025 <- readRDS("../data/processed/after_processed.rds")
after_final <- left_join(spillover_all_2025, full_2025, by = "census_id")
saveRDS(after_final, "../data/processed/after_final.rds")
```


```{r}
panel_data <- bind_rows(before_final %>% mutate(year = 2015, post = 0),
                        after_final %>% mutate(year = 2024, post = 1))
saveRDS(panel_data, "../data/processed/both_years.rds")
```


## PLot sigma 
```{r}
library(ggplot2)
ggplot() +
  geom_sf(data = census_sections) +
  geom_sf(data = listings2015_sf, alpha = 0.5) +
  geom_sf(data = st_buffer(listings2015_sf, dist = 300), fill = "red", alpha = 0.1) +
  ggtitle("Visualizing Sigma = 300m")

```
```{r}
panel_data <- bind_rows(census_analysis_before %>% mutate(year = 2015, post = 0),
                        census_analysis_after %>% mutate(year = 2024, post = 1))
saveRDS(panel_data, "../data/processed/both_years.rds")
```



