---
title: "spatial"
author: "Laura Lundbye"
date: "2025-05-29"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(dplyr)
library(tidyr)
library(ggplot2)
library(brms)
```

#Fititng and running the models

```{r}
setwd("C:/Users/laura/Desktop/uni/cultural_spacial")
```

```{r}
data <- read.csv("both_years.csv")
```

## Initial preprocessing
First, I preprocess the data further to fit with modelling requirements. This includes creating a categorical before and after column form the year column (0 for 2015 and 1 for 2024).
Next, I investigate the distribution of the education level distribution for both years to see how I want to group the education levels.
I remove edu_6 as its most likely a fluke category, as it's very small for 2015 and contains only 0s for 2024.
I also remove the geometry column as it's no longer needed and causes some issues due to its shape.

```{r}
#creating before and after column and making sure it's categorical
data$before_after <- ifelse(data$year == 2015, 0,
                            ifelse(data$year == 2024, 1, NA))
data <- data %>%
  mutate(before_after = as.factor(before_after))

data$geometry <- NULL
data$pct_edu_6 <- NULL

#turning NAs into 0s
data <- data %>%
  mutate(airbnb_per_1000 = replace_na(airbnb_per_1000, 0))

#renaming column
data <- data %>%
  rename(disposable_income = Import_Euros)

```

## Plotting
In the next two chunks, i plot the education data as two simple categorical histograms

```{r}
#plotting education for 2015
data_2015 <- data %>% filter(year == 2015)

#reshape 2015 data to long format
edu_2015_long <- data_2015 %>%
  select(starts_with("pct_edu_"), year) %>%
  pivot_longer(cols = starts_with("pct_edu_"),
               names_to = "education_level",
               values_to = "value")

#plot for 2015
ggplot(edu_2015_long, aes(x = education_level, y = value)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  labs(title = "Education Levels in 2015",
       x = "Education Level",
       y = "Percentage") +
  theme_minimal()
```

```{r}
#plotting education for 2024
data_2024 <- data %>% filter(year == 2024)

#reshaping
edu_2024_long <- data_2024 %>%
  select(starts_with("pct_edu_"), year) %>%
  pivot_longer(cols = starts_with("pct_edu_"),
               names_to = "education_level",
               values_to = "value")

#plot for 2024
ggplot(edu_2024_long, aes(x = education_level, y = value)) +
  geom_bar(stat = "identity", fill = "darkgreen") +
  labs(title = "Education Levels in 2024",
       x = "Education Level",
       y = "Percentage") +
  theme_minimal()

```
## Grouping education
Based on the results in the histograms, I choose to make the cut between level 3 and 4 due to concentration.
I then factor census_id for modelling

```{r}
data <- data %>%
  mutate(percentage_of_edu_high = pct_edu_4 + pct_edu_5)

data$census_id <- as.factor(data$census_id)
```

## Normalizing variables
Next I normalize the spill-over effect values at 300 m, the Airbnb density per 1000 residents counts, and the disposable income counts into z-scores.
This helps my models converge and makes for better modelling all around

```{r}
#normalizing conitnuys variables
data <- data %>%
  mutate(
    kernel_spill_300_z = as.numeric(scale(kernel_spill_300)),
    airbnb_per_1000_z = as.numeric(scale(airbnb_per_1000)),
    disposable_income_z = as.numeric(scale(disposable_income))
  )
```

## Education model
Using brms, I fit my frist model. This one predicts the percentage of residents with high education in various censuses by the following predictors: the year measure, Airbnb density, the interaction of the two, the spill-over values at 300 m, and random effects for every census.
As the education percentages are fairly spanning 0.2 - 0.8, I feel comfortable using a Gaussian family instead of a beta family or another family geared toward data between 0 and 1. Bulk Effective Samples Size (ESS) was fairly low for this model, so I cranked the chains and iterations up by quite a bit. This lets them reach a sufficient count for every beta value, as NUTS sampling can usually deal with a minimum of 1000 (check the Bulk_ESS counts in the summary for exact numbers - this model scores a lowest count of around 1700)
The posterios are plotted and the model output is summed and printed

```{r}
model_edu <- brm(
  formula = percentage_of_edu_high ~ before_after * airbnb_per_1000_z + kernel_spill_300_z + ( 1 | census_id),
  data = data,
  family = gaussian(),
  chains = 4,
  cores = 4,
  iter = 3000,
  warmup = 1500,
  seed = 123,
  control = list(max_treedepth = 15)
)
```

```{r}
summary(model_edu)
plot(model_edu)
```

## Disposable income model
Now I create my second model. This one predicts disposable income by the same predictors as the first model. This model scores bulk ESS scores of 2000+.
```{r}
model_income <- brm(
  formula = disposable_income_z ~ before_after * airbnb_per_1000_z + kernel_spill_300_z + (1 | census_id),
  data = data,
  family = gaussian(),
  chains = 4,
  cores = 4,
  iter = 3000,
  warmup = 1000,
  seed = 123
)

```

```{r}
summary(model_income)
plot(model_income)
```
##Visualizing the data

###Visualizing higher education percentage
Here i plot the percentage of residents with higher education by Airbnb density for every census for both years.

```{r}
#renaming the before_after measure
data$before_after <- factor(data$before_after, levels = c(0, 1), labels = c("2015", "2024"))

ggplot(data, aes(x = airbnb_per_1000, y = percentage_of_edu_high, color = before_after)) +
  geom_point(alpha = 0.5) +
  geom_line(aes(group = census_id), color = "gray60", alpha = 0.2) +  # connect points from same census tract
  geom_smooth(method = "lm", se = FALSE) +  # regression lines per year
  labs(
    title = "Relationship between Airbnb Density and High Education Levels",
    x = "Airbnb Listings per 1000 Residents",
    y = "Percentage with Higher Education",
    color = "Year"
  ) +
  ylim(0, 1) +
  theme_minimal()

```

###Visualizing disposable income
Finally, I create a plot like the one above but for disposable income instead of higher education.

```{r}
#plotting income

data$before_after <- factor(data$before_after, levels = c(0, 1), labels = c("2015", "2021"))

ggplot(data, aes(x = airbnb_per_1000, y = disposable_income, color = before_after)) +
  geom_point(alpha = 0.5) +
  geom_line(aes(group = census_id), color = "gray60", alpha = 0.2) +  # connect points from same census tract
  geom_smooth(method = "lm", se = FALSE) +  # regression lines per year
  labs(
    title = "Relationship between Airbnb Density and Disposable Income",
    x = "Airbnb Listings per 1000 Residents",
    y = "Disposable Income",
    color = "Year"
  ) +
  theme_minimal()
```


